---
title: "Data Preparation for AMR in SSI Systematic Review"
author: "Emmanuel Korir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
# install-if-missing helper
install_if_missing <- function(pkgs) {
  for (p in pkgs) {
    if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
  }
}

# ensure rprojroot is available
install_if_missing(c("rprojroot"))

# load libraries (safe)
library(rprojroot)

# knitr options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# try to find RStudio project root; if not found, fall back to current working dir
root_dir <- tryCatch(
  rprojroot::find_rstudio_project_file(),
  error = function(e) {
    message("rprojroot couldn't find an RStudio project file; using current working directory.")
    NULL
  }
)

if (!is.null(root_dir)) {
  knitr::opts_knit$set(root.dir = root_dir)
} else {
  # optionally set to getwd() explicitly
  knitr::opts_knit$set(root.dir = getwd())
}
```

## Load Libraries and Data

```{r load-libraries-and-data}

install_if_missing <- function (packages) {
    for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg)
    }
  }
}

# List of packages you want
packages <- c("tidyverse", "dplyr", "gtsummary", "meta", "stringr", "maps", "countrycode", "mapproj")

# Install if missing
install_if_missing(packages)

library(tidyverse)
library(dplyr)
library(gtsummary) 
library(meta)
library(stringr)
library(maps)
library(countrycode)
library(mapproj)

# Load the clean data you saved in the last step
df_final <- readRDS("../data/processed_data/cleaned_amr_ssi_data.rds")
```

### List column names
```{r}
glimpse(df_final)

## colnames(df_final)/str(df_final) bot lists column names just as glimpse()
```


## 1. Study Characteristics

```{r create-table-1}
# Select the key columns you want to summarize
table1_data <- df_final %>%
  select(
    `Publication Year` = year_of_publication,
    `Country` = country_countries,
    `Study Design` = study_design,
    `Total Sample Size` = n_total_sample,
    `SSI Incidence Rate (%)` = prop_ssi_incidence,
    `Percent Female` = prop_female
  )

# Create the summary table
table1_data %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    missing_text = "Not Reported"
  ) %>%
  add_overall() %>%
  modify_caption("Table 1: Characteristics of Included Studies")
```

## 2. Burden of SSI Incidence

Meta analysis

```{r}
meta_data <- df_final %>%
  select(study_id, n_ssi, n_procedures) %>%
  # A study needs both the number of SSIs and the total number of procedures
  filter(!is.na(n_ssi) & !is.na(n_procedures))

# Perform the meta-analysis of proportions (incidences)
meta_prop <- metaprop(
  event = n_ssi,
  n = n_procedures,
  studlab = study_id,
  data = meta_data,
  sm = "PFT", # Use Freeman-Tukey double arcsine transformation for proportions
  method.tau = "REML" # A standard method for estimating heterogeneity
)

# Print the summary of the meta-analysis
summary(meta_prop)

# Create a forest plot - the classic visualization for a meta-analysis
forest(meta_prop)
```

Subgroup analysis can be added later

```{r meta-analysis-subgroup}
meta_subgroup_data <- df_final %>%
  select(study_id, n_ssi, n_procedures, country_countries) %>%
  filter(!is.na(n_ssi) & !is.na(n_procedures) & !is.na(country_countries)) %>%
  add_count(country_countries) %>%
  filter(n > 1) 

# Check if there's enough data to proceed
if(nrow(meta_subgroup_data) > 0) {
  # Perform the meta-analysis with subgroups
  meta_prop_subgroup <- metaprop(
    event = n_ssi,
    n = n_procedures,
    studlab = study_id,
    data = meta_subgroup_data,
    sm = "PFT",
    method.tau = "REML",
    byvar = country_countries # This creates the subgroup analysis
  )

    # Print the summary and create a forest plot
  summary(meta_prop_subgroup)
  forest(meta_prop_subgroup)
} else {
  print("Not enough recurring data for a meaningful subgroup analysis by country.")
}

### some studies are repeated, would delete them later, fix this error

####narrative for the paper: The overall pooled incidence of SSIs across all studies was [X.XX%] (95% CI: Y.YY - Z.ZZ), as determined by a random-effects meta-analysis. Substantial heterogeneity was observed (I² = XX%, p < 0.001). When stratified by country, the pooled incidence was highest in [Country A] at [A.A%] and lowest in [Country B] at [B.B%] (Figure X).
```
#### Geographic bubble map

```{r geographic-bubble-map, eval=FALSE}

df_final_cleaned <- df_final %>%
  mutate(
    # Create a new column specifically for mapping
    country_for_map = case_when(
      # Rule 1: Standardize the name for DRC
      country_countries == "Democratic Republic of Congo" ~ "Democratic Republic of the Congo",
      
      # Rule 2: Identify and exclude the multi-country study by setting it to NA.
      # str_detect() finds any row containing this specific text pattern.
      str_detect(country_countries, "Benin, Ghana, Nigeria") ~ NA_character_,
      
      # Rule 3: For all other countries that are fine, keep their original names.
      TRUE ~ country_countries
    )
  )

# Optional: You can print a message to confirm which studies were excluded from the map
multi_country_studies <- df_final_cleaned %>%
  filter(is.na(country_for_map)) %>%
  select(study_id, country_countries)

if(nrow(multi_country_studies) > 0){
  message("The following studies were identified as multi-country and will be excluded from the map:")
  print(multi_country_studies)
}

# Step 1: Summarize the CLEANED data by country
country_summary <- df_final_cleaned %>%
  # Filter out rows with no SSI data or the multi-country studies we set to NA
  filter(!is.na(prop_ssi_incidence) & !is.na(country_for_map)) %>%
  group_by(country_for_map) %>% # Use the new clean column
  summarise(
    mean_ssi_rate = mean(prop_ssi_incidence, na.rm = TRUE),
    n_studies = n()
  ) %>%
  ungroup()

# Step 2: Get world country centroids (center points)
world_map_data <- map_data("world")
country_centroids <- world_map_data %>%
  group_by(region) %>%
  summarise(long = mean(long), lat = mean(lat))

# Step 3: Join your summary data with the coordinate data
map_plot_data <- left_join(country_summary, country_centroids, by = c("country_for_map" = "region"))

# --- PART 4: Create the Plot with a Dynamic Caption (The New Part) ---

# First, create the text for our caption
# This automatically counts the number of excluded studies.
caption_text <- paste0("Note: ", nrow(multi_country_studies), " multi-country study/studies were excluded from this map.")

# Now, create the plot and add the caption
ggplot() +
  geom_polygon(data = world_map_data, aes(x = long, y = lat, group = group), fill = "gray90", color = "white") +
  geom_point(data = map_plot_data, aes(x = long, y = lat, size = n_studies, color = mean_ssi_rate), alpha = 0.8) +
  scale_size_continuous(range = c(4, 15), name = "Number of Studies") +
  scale_color_viridis_c(option = "plasma", name = "Mean SSI Rate (%)") +
  coord_map(xlim = c(-20, 55), ylim = c(-35, 38)) +
  
  # This is the line that adds the text TO THE PLOT
  labs(
    title = "Geographic Distribution and Burden of SSIs in Africa",
    subtitle = "Bubble size represents number of studies; color represents mean SSI incidence rate",
    x = NULL, y = NULL,
    caption = caption_text # ADDED THIS LINE
  ) +
  
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic") # Style the caption
  )

```


### Stratified SSI Incidence

```{r stratified-ssi-incidence, eval=FALSE}
# Filter for studies that report specialty and have the necessary numbers
meta_specialty_data <- df_final %>%
  select(study_id, n_ssi, n_procedures, surgical_speciality) %>%
  filter(!is.na(n_ssi) & !is.na(n_procedures) & !is.na(surgical_speciality)) %>%
  add_count(surgical_speciality) %>%
  filter(n > 1) # Subgroup analysis requires at least 2 studies per group

if(nrow(meta_specialty_data) > 0) {
  meta_prop_specialty <- metaprop(
    event = n_ssi,
    n = n_procedures,
    studlab = study_id,
    data = meta_specialty_data,
    sm = "PFT",
    byvar = surgical_speciality # The only line that changes!
  )
  
  forest(meta_prop_specialty)
} else {
  print("Not enough data for a meaningful subgroup analysis by surgical specialty.")
}

```


## 3. Challenges

### 3a. Aetiological Agents

List of most common pathogens in ECSA

```{r}
# Reshape the data to make it easy to count pathogens
pathogen_long <- df_final %>%
  select(study_id, pathogen_1_name_name_of_the_most_common_isolated_pathogen, pathogen_2_name_name_of_the_2nd_most_common_pathogen, pathogen_3_name_name_of_the_3rd_most_common_isolated_pathogen) %>%
  pivot_longer(
    cols = starts_with("pathogen"),
    names_to = "pathogen_rank",
    values_to = "pathogen_name"
  ) %>%
  filter(!is.na(pathogen_name)) # Remove rows where there was no pathogen listed

# Now we can easily count the most frequently reported pathogens
pathogen_counts <- pathogen_long %>%
  count(pathogen_name, sort = TRUE, name = "count")

# Visualize the top 10 most common pathogens
pathogen_counts %>%
  top_n(10, count) %>%
  ggplot(aes(x = reorder(pathogen_name, count), y = count)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  labs(
    title = "Top 10 Most Frequently Reported Pathogens in SSIs",
    x = "Pathogen",
    y = "Number of Times Reported (as a Top 3 Pathogen)"
  ) +
  theme_minimal()
```

### 3b. AMR

Analysis of resistance patterns for the key pathogens

```{r parse-and-analyze-resistance}

# Step 1: Reshape data with explicit renaming for clarity and robustness
# We select and rename the long original columns to a simple, predictable pattern 
# like 'pathogen_name_1', 'resistance_1', etc.
pathogens_long <- df_final %>%
  select(
    study_id,
    pathogen_name_1 = `pathogen_1_name_name_of_the_most_common_isolated_pathogen`,
    resistance_1 = `pathogen_1_resistance_list_antibiotic_resistance`,
    pathogen_name_2 = `pathogen_2_name_name_of_the_2nd_most_common_pathogen`,
    resistance_2 = `pathogen_2_resistance_list_antibiotic_resistance_percent`, # Note the slightly different name here
    pathogen_name_3 = `pathogen_3_name_name_of_the_3rd_most_common_isolated_pathogen`,
    resistance_3 = `pathogen_3_resistance_list_antibiotic_resistance`
  ) %>%
  # NOW, pivot_longer can work on our clean, simple names
  pivot_longer(
    cols = -study_id,
    names_to = c(".value", "rank"), # This creates new columns from the first part of the name
    names_pattern = "([a-z_]+)_(\\d)" # Splits "pathogen_name_1" into "pathogen_name" and "1"
  )

# At this point, pathogens_long has columns named `study_id`, `rank`, `pathogen_name`, and `resistance`.
# The error is now fixed because the `pathogen_name` column exists.

# Step 2: Parse the resistance text string for each pathogen
# (This part of the code remains the same as it operates on the now-correctly-shaped data)
resistance_tidy <- pathogens_long %>%
  # Filter out rows where there is no pathogen name or no resistance data
  filter(!is.na(pathogen_name) & !is.na(resistance) & resistance != "") %>%
  # Rename the 'resistance' column which contains the raw text string for clarity
  rename(resistance_data = resistance) %>%
  # Split the comma-separated list into separate rows
  separate_rows(resistance_data, sep = ",\\s*") %>%
  # Extract the antibiotic name and the resistance percentage using regex
  mutate(
    antibiotic_name = str_trim(str_extract(resistance_data, "^[A-Za-z\\s\\-]+")),
    resistance_rate = as.numeric(str_extract(resistance_data, "\\d+\\.?\\d*"))
  ) %>%
  # Clean up and keep only valid rows
  filter(!is.na(antibiotic_name) & !is.na(resistance_rate)) %>%
  select(study_id, pathogen_name, antibiotic_name, resistance_rate)

# Step 3: Analyze and Visualize the Resistance Data
# (This part of the code also remains the same)
# Let's analyze resistance for the top 3 most common pathogens
top_pathogens <- c("Staphylococcus aureus", "Escherichia coli", "Klebsiella pneumoniae")

resistance_summary <- resistance_tidy %>%
  filter(pathogen_name %in% top_pathogens) %>%
  group_by(pathogen_name, antibiotic_name) %>%
  summarise(
    n_studies_reported = n(),
    mean_resistance = mean(resistance_rate, na.rm = TRUE),
    median_resistance = median(resistance_rate, na.rm = TRUE),
    .groups = 'drop' # Recommended to add .groups = 'drop' after summarise
  ) %>%
  filter(n_studies_reported >= 2) %>% 
  arrange(pathogen_name, desc(mean_resistance))

# Print the summary table
print(resistance_summary)

# Create a faceted plot to show resistance for each top pathogen
resistance_summary %>%
  ggplot(aes(x = mean_resistance, y = reorder(antibiotic_name, mean_resistance), fill = pathogen_name)) +
  geom_col() +
  geom_text(aes(label = paste0(round(mean_resistance), "%")), hjust = -0.2, size = 3) +
  facet_wrap(~pathogen_name, scales = "free_y") +
  labs(
    title = "Mean Antimicrobial Resistance Rates for Top Pathogens",
    x = "Mean Resistance Rate (%)",
    y = "Antibiotic"
  ) +
  theme_bw() +
  theme(legend.position = "none") +
  xlim(0, 110)
```
Narrative for the paper:
The most frequently isolated pathogens were Staphylococcus aureus, Escherichia coli, and Klebsiella pneumoniae. Resistance patterns for these organisms were extracted (Table X). Staphylococcus aureus demonstrated high mean resistance to [Ampicillin] ([XX.X%]) and [Penicillin] ([YY.Y%]), but remained susceptible to [Vancomycin] ([Z.Z%]). Among Gram-negative isolates, E. coli showed alarming resistance rates to third-generation cephalosporins like [Ceftriaxone] ([AA.A%]). Figure Y visualizes the resistance profiles for the most commonly reported pathogen-antibiotic combinations.

### 3c. Clinical and Economic Burden

Summary of morbidity, mortality, and costs
```{r burden-analysis}
burden_data <- df_final %>%
  select(
    `Additional Hospital Stay (Days)` = morbidity_add_hosp_stay_days,
    `Re-operation Rate (%)` = prop_reoperation_rate,
    `Readmission Rate (%)` = prop_readmission_rate,
    `SSI-Attributable Mortality Rate (%)` = prop_mortality_ssi_attributable
  )

# Create a summary table of these outcomes
burden_data %>%
  tbl_summary(
    statistic = all_continuous() ~ "{mean} ({sd}) | {median} ({p25}, {p75})",
    label = list(
      `Additional Hospital Stay (Days)` ~ "Additional Hospital Stay (Mean Days)",
      `Re-operation Rate (%)` ~ "Re-operation (Mean %)",
      `Readmission Rate (%)` ~ "Readmission (Mean %)",
      `SSI-Attributable Mortality Rate (%)` ~ "SSI-Attributable Mortality (Mean %)"
    ),
    missing_text = "Not Reported"
  ) %>%
  modify_caption("Table X: Clinical and Economic Burden Associated with SSIs")
```
Narrative: 
"The clinical burden of SSIs was substantial. Across studies that reported it, SSIs were associated with a mean additional hospital stay of [X.X] days. The mean re-operation rate was [Y.Y%], and the mean SSI-attributable mortality was [Z.Z%] (Table X)."

### 3d. WHO critical important antibiotics

```{r who-amr-analysis, eval=FALSE}
# First, run the 'parse-and-analyze-resistance' chunk to create 'resistance_tidy'

# Define a list of WHO "Critically Important Antimicrobials" (this is an example list, expand as needed)
# Source: WHO AWaRe classification
who_critical_abx <- c(
  # Cephalosporins (3rd gen and higher)
  "Ceftriaxone", "Cefotaxime", "Ceftazidime", "Cefepime",
  # Carbapenems
  "Meropenem", "Imipenem", "Ertapenem",
  # Glycopeptides
  "Vancomycin",
  # Fluoroquinolones
  "Ciprofloxacin", "Levofloxacin"
)

# Analyze resistance specifically to these antibiotics
who_resistance_summary <- resistance_tidy %>%
  # Filter for only the antibiotics in our critical list
  filter(antibiotic_name %in% who_critical_abx) %>%
  group_by(pathogen_name, antibiotic_name) %>%
  summarise(
    n_studies = n(),
    mean_resistance = mean(resistance_rate),
    .groups = 'drop'
  ) %>%
  arrange(pathogen_name, desc(mean_resistance))

# Print the table
print(who_resistance_summary)

# Visualize it
who_resistance_summary %>%
  ggplot(aes(x = mean_resistance, y = reorder(antibiotic_name, mean_resistance), fill = pathogen_name)) +
  geom_col(position = "dodge") +
  labs(
    title = "Resistance to WHO Critically Important Antimicrobials",
    x = "Mean Resistance Rate (%)",
    y = "Antibiotic",
    fill = "Pathogen"
  ) +
  theme_minimal()
```

Narrative:
Resistance to WHO Critically Important Antimicrobials was widespread. For example, the mean resistance of E. coli to Ceftriaxone was [XX.X%] across [N] studies. Alarming rates of resistance to carbapenems were noted for Pseudomonas aeruginosa (Figure Z).
## 4. Innovation and Straegies

### 4a. Reported Interventions and Innovations

### 4b. Reported Drivers of AMR and Gaps

```{r}
# We will use the same text analysis technique as for the resistance data

# 4a. Interventions and Innovations Described
interventions_counts <- df_final %>%
  select(interventions_innovations_described) %>%
  filter(!is.na(interventions_innovations_described) & interventions_innovations_described != "") %>%
  separate_rows(interventions_innovations_described, sep = ",\\s*|;\\s*") %>%
  count(interventions_innovations_described, sort = TRUE, name = "number_of_studies")

# 4b. Reported Drivers of AMR
drivers_counts <- df_final %>%
  select(reported_drivers_of_amr) %>%
  filter(!is.na(reported_drivers_of_amr) & reported_drivers_of_amr != "") %>%
  separate_rows(reported_drivers_of_amr, sep = ",\\s*|;\\s*") %>%
  count(reported_drivers_of_amr, sort = TRUE, name = "number_of_studies")

# 4c. Gaps Identified by Authors
gaps_counts <- df_final %>%
  select(gaps_identified_by_authors) %>%
  filter(!is.na(gaps_identified_by_authors) & gaps_identified_by_authors != "") %>%
  separate_rows(gaps_identified_by_authors, sep = ",\\s*|;\\s*") %>%
  count(gaps_identified_by_authors, sort = TRUE, name = "number_of_studies")

# Display these as tables using gtsummary or gt
# Example for Interventions
interventions_counts %>%
  gt::gt() %>%
  gt::cols_label(
    interventions_innovations_described = "Intervention or Innovation",
    number_of_studies = "Number of Studies Mentioning"
  ) %>%
  gt::tab_header("Table Y: Reported Strategies and Innovations to Combat AMR in SSIs")

# Display drivers
drivers_counts %>%
  gt::gt() %>%
  gt::cols_label(
    reported_drivers_of_amr = "Reported Driver of AMR",
    number_of_studies = "Number of Studies Mentioning"
  ) %>%
  gt::tab_header("Table Z: Perceived Drivers of AMR")
```

Narrative for the paper:
Several studies described interventions aimed at combating SSIs and AMR. The most frequently mentioned strategies included [e.g., 'Enhanced hand hygiene programs'] (N=X studies), ['Implementation of antibiotic stewardship'] (N=Y studies), and ['Pre-operative screening for MRSA'] (N=Z studies) (Table Y). Authors of the included studies identified several drivers of AMR, with ['Over-the-counter antibiotic sales'] and ['Poor infection control'] being the most cited (Table Z). Key gaps identified for future research included the need for ['Robust surveillance systems'] and ['Point-of-care diagnostics'].

## 5. Morbidity, Hospital Burden, and Economic Costs

```{r}
# Part A: Quantitative Burden Metrics
# We are cleaning up the mortality column first, assuming it might have '%' signs
burden_quantitative <- df_final %>%
  mutate(
    mortality_90_day = as.numeric(str_remove(`mortality_90_day_post_op_percent`, "%"))
  ) %>%
  select(
    `Additional Hospital Stay (Days)` = morbidity_add_hosp_stay_days,
    `Total Length of Stay (Days)` = hospital_burden_total_length_of_stay_days,
    `Re-operation Rate (%)` = prop_reoperation_rate,
    `Readmission Rate (%)` = prop_readmission_rate,
    `SSI-Attributable Mortality Rate (%)` = prop_mortality_ssi_attributable,
    `30-Day Post-op Mortality (%)` = prop_mortality_30day_post_op,
    `90-Day Post-op Mortality (%)` = mortality_90_day
  )

# Create the summary table
burden_quantitative %>%
  tbl_summary(
    statistic = all_continuous() ~ "{mean} ({sd})",
    digits = all_continuous() ~ 1,
    missing_text = "Not Reported"
  ) %>%
  modify_caption("Table X: Clinical Burden and Mortality Associated with SSIs")

# Part B: Qualitative Economic Cost Analysis
economic_costs <- df_final %>%
  select(study_id, `Direct Costs` = economic_direct_costs, `Indirect Costs` = economic_indirect_costs) %>%
  pivot_longer(cols = -study_id, names_to = "cost_type", values_to = "cost_description") %>%
  filter(!is.na(cost_description) & cost_description != "") %>%
  separate_rows(cost_description, sep = ",\\s*|;\\s*") %>%
  count(cost_type, cost_description, sort = TRUE)

# Display as a table
economic_costs %>%
  gt::gt() %>%
  gt::tab_header("Table Y: Reported Economic Costs of SSIs")
```



